---
title: 'Zero hour: 9:00 a.m.'
author: "Sean"
date: "5/11/2020"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	message = FALSE,
	warning = FALSE
)
```


## Looking for the Right Stuff

Space is an excellent area for datascience projects, as the Space Operations and Astronomy fields have been collecting vast amounts of data for decades.  In fact, there's so much data floating around in digital space that it was difficult to decide where to begin.  NASA alone has a repository of [~10,000 datasets](https://data.nasa.gov/).

So to narrow my search, I decided to look at the foundational, and most explosive, aspect of Space Exploration - Rocketry.  In this search I found Jonathan McDowell's suborbital and orbital launch database at [planet4589.org](https://planet4589.org/space/lvdb/index.html).  This expansive dataset includes listing of 72,920 individual launches; cataloging information including launch date, vehicle, estimated apogeee, organization, and much more.

## Houston we have a Problem

Any dataset, especially one that is manually input over a long period of time, will have some inherent quirks that are difficult for computers to parse and analyze.  An example from this dataset is questionmarks added to the end of an Apogee measure (max altitude of the rocket in miles).   McDowell's notes explain that Apogee questionmarks are estimations with up to 20% error.  We humans can now understand that an apogee of "100?" is an approximate number, but a computer would struggle to even recognize this as a number.  So to begin any analysis, we need to "clean" these values.

Here's NASA solving the same square-peg, round-hole problem (just in a slightly more critical situation).

<iframe width="560" height="315" src="https://www.youtube.com/embed/ry55--J4_VQ" frameborder="0" allow="accelerometer; autoplay; encrypted-media; gyroscope; picture-in-picture" allowfullscreen></iframe>

To succesfully analyze our data, we need change the data's format into something understandable by our computer language (in this case R).  With McDowell's detailed notes on his various files and header information we can use our understanding of the data to manipulate it into a form that our computer can understand.  After converting our file formats, removing some characters, and filling in a few missing details (only 6 missing organizations of 72k entries) we're ready to go.

## One Small Step

Now that the data is relatively clean we can begin our analysis.  We'll start off simply by plotting all launch attempts through history.  This gives us a general view of the predominance of rocketry through time.

```{r}
library(tidyverse)
library(ggplot2)
library(gganimate)
library(gifski)
library(pins)
library(zoo)
library(treemapify)
library(networkD3)
library(ggforce)

# See Jupyter Notebook file for cleaning
df0 <- read_csv('C:/Users/ssq10/Documents/DataBrahs/5 May - Space Race/launch_test.csv') %>%
  mutate(Agency = gsub("(.*?)(/.*)", "\\1", Agency),
         Agency = gsub("\\?", "", Agency),
         Agency = case_when(
           LV_Type == 'Burya' ~ 'SU', # 5 Burya Rockets Were Unlabeled by Org
           Launch_Tag == '2019-S81' ~ 'USAF', # 1 USAF Ballistic Missile Was Unlabeled
           T ~ Agency
         ))

# Organization and Country lists
df_org <- read_tsv('C:/Users/ssq10/Documents/DataBrahs/5 May - Space Race/orgs.tsv') #tsv created from original Orgs TAR File
df_cntry <- df_org %>% select(`#Code`,StateCode, ShortName, ShortEName) %>%
  distinct()
colnames(df_cntry) <- c('Agency','StateCode','AgencyName', 'StateName')

Europe <- c('Germany', 'Germany (Reich)',
                     'France','Denmark','ELDO','ESA',
                     'ESRO','Italy','Norway','Spain',
                     'Sweden','Switzerland', 'UK')
Asia <- c('India','Pakistan','North Korea',
                     'Taiwan','China','Indonesia',
                     'India','South Korea','Japan')
MidEast <- c('Egypt','Iran','Israel',
                     'Iraq','Yemen','Libya',
                     'Syria')
Americas <- c('Brazil','Argentina', 'Canada', 'Cayman Is.')
Oceania <- c('Australia','New Zealand')

df_codes <- df_cntry %>%
  filter(Agency == StateCode) %>%
  select(StateCode, StateName)


#Building Initial Dataframes
df0 <- left_join(df0,df_cntry) %>%
  drop_na(StateCode) # Bunch of random NA Rows

df0a <- df0 %>%
  mutate(Year = substring(Launch_Date, 1, 4), # First portion of mutate to convert time to columns
         Month = substring(Launch_Date, 6, 8),
         Day = substring(Launch_Date, 10, 11),
         Hour = substring(Launch_Date, 13, 14),
         Minute = substring(Launch_Date, 15, 16),
         Decade = paste(substring(Year, 1, 3), "0", sep = ''), # Add Decade Variable 
         Apogee = as.numeric(gsub("\\?", "", Apogee))) # Remove Question Marks from Apogee 

# As.date(paste(year,mo,sep="-"),format ="%Y-%b
launch_by_agency <- function(data){
  df2 <- data %>%  #Group By Agency and Launch Year
    mutate(StateCode = case_when(
      Decade < '1990' & StateCode == 'RU' ~ 'SU',
      Decade >= '1990' & StateCode == 'SU' ~ 'RU',
      T ~ StateCode
    )) %>%
    group_by(AgencyName, StateCode, Decade, Year) %>%
    summarize(yr_launch = n())
  
  return(df2)
}

df_ann_join <- launch_by_agency(df0a)

df_mrn <- df_ann_join %>%
  filter(AgencyName == 'MRN') %>%
  mutate(Year = as.numeric(Year))

df_ann_plt <- df_ann_join %>%
  group_by(Year) %>%
  summarize(yr_launch = sum(yr_launch)) %>%
  mutate(Year = as.numeric(Year))

ggplot() +
  geom_path(data = df_ann_plt, aes(x = Year, y = yr_launch), color = 'black', size = 1.5) +
  labs(title = 'All Rocket Launches by Year', x = 'Year', y ='Launches') +
  theme_bw()
```

We see here a sharp increase in the usage of rockets during their initial development and then a steady decline to a relative steady state.  In the plot we can see that the peak rocket usage was during the 1970s.

But what's the major factor that caused this spike?  Was it the height of the cold war?  A race to get more satellites?  ICBM testing?  To get a better understanding of what was driving this increase we broke our data down to individual organizations.  What we found was something rather interesting.

```{r}
ggplot() +
  geom_path(data = df_ann_plt, aes(x = Year, y = yr_launch), color = 'black', size = 1.25) +
  geom_bar(data = df_mrn, aes(x = Year, y = yr_launch, fill = AgencyName), stat = 'identity') +
  labs(title = 'All Rocket Launches by Year', x = 'Year', y ='Launches', fill = 'Top Organization') +
  theme_bw() +
  theme(legend.position = 'bottom')
```

Here we see the massive increase in launches during the 70s is driven by one organization.  A little known US organization called the Meteorological Rocket Network, which comprised of members from different military, meteorological, and scientific agencies.  Their aim was to measure atmospheric conditions in the areas between weather balloons and satellites.  What's pretty shocking is that MRN accounts for 33% of the world's total rocket launches, but doesn't even have a wikipedia page.  Below are some more details on their launches.

### By the Numbers

```{r}
mrn <- df0a %>% 
  filter(AgencyName == 'MRN')
```
Meteorological Rocket Network:

Total Launches : ``r nrow(mrn)``

Total Launches >100 Apogee : ``r nrow(filter(mrn, Apogee > 100))``

Mean Launch Apogee : ``r mean(mrn$Apogee)``

Max Launch Apogee : ``r max(mrn$Apogee)``

Launches by Vehicle Type :
```{r} 
table(mrn$LV_Type)
```


## Bigger is Better

Let's focus the remainder of our analysis on the major rocket launches, or launches that achieved an estimated apogee of at least 100 miles.  Of all launches, 24,261 meet this criteria.  Our first look will be launches divided by the US, USSR/Russia, and the continents.  This will let us see patterns by regions throughout the years.

```{r}
df1 <- df0a %>%
  filter(Apogee >= 100)

launch_by_year <- function(data){
  df2 <- launch_by_agency(data)
  
  df_join <- expand.grid(Year = as.character(seq(min(df2$Year),
                                                 max(df2$Year), by =1)),
                         AgencyName = unique(df2$AgencyName)) %>%
    left_join(df2, by = c("Year","AgencyName")) %>%
    mutate(yr_launch = case_when(
      is.na(yr_launch) == T ~ as.numeric(0),
      T ~ as.numeric(yr_launch))) %>%
    group_by(AgencyName, Decade) %>%
    mutate(launches = cumsum(yr_launch)) %>%  
    mutate(grtr_1 = case_when(
      launches > 1 ~ 1,
      T ~ 0
    )) %>%
    group_by(AgencyName, grtr_1) %>%
    mutate(count_1 = seq(n())) %>%
    ungroup() %>%
    filter(grtr_1 == 1)
  
  return(df_join)}

df_join <- launch_by_year(df1)

df_decade <- df_join %>%
  group_by(AgencyName, Decade) %>%
  filter(row_number() == n()) %>%
  left_join(df_cntry) %>%
  distinct(AgencyName, .keep_all = T) %>%
  select(-StateName) %>% 
  mutate(StateCode = case_when(
    Decade < '1990' & StateCode == 'RU' ~ 'SU',
    Decade >= '1990' & StateCode == 'SU' ~ 'RU',
    T ~ StateCode
  )) %>%
  left_join(df_codes, by =c ('StateCode', 'StateCode')) %>%
  select(AgencyName, Decade, launches, StateName)


df_dcd_cntry <- df_decade %>%
  group_by(Decade, StateName) %>%
  summarize(launches = sum(launches)) %>%
  mutate(main = case_when(
    StateName == 'USA' ~ 'USA',
    StateName %in% c('USSR','Russia') ~ 'Russia/USSR',
    StateName %in% Europe  ~ 'Europe (Not Russia)',
    StateName %in% MidEast ~ 'Mid-East',
    StateName %in% Asia ~ 'Asia',
    StateName %in% Americas ~ 'Americas (Not USA)',
    StateName %in% Oceania ~ 'Oceania',
    T ~ StateName
  )) %>%
  filter(Decade != '2020')

ggplot(data=df_dcd_cntry, aes(x = Decade, y = launches, fill = main), color = main) + 
  geom_bar(position="stack", stat="identity") +
  scale_fill_hue(l=50, h.start = 250) +
  scale_color_hue(l=50, h.start = 250) +
  labs(title = 'Major Rocket Launches by Decade', y = '# of Launches', x = '',
       caption = 'Major Rocket Launch is any launch >100 Apogee',
       fill = 'Area') + theme(legend.position="right") +
  theme_light()
```

Now we see that the most active period for major launches has shifted left from our total launches to the 1960s.  This makes sense as it was peak of the space race between the US and the Soviet Union.  

Next we'll see what the proportion of those launches were between the regions.

```{r}
ggplot(data=df_dcd_cntry, aes(x = Decade, y = launches, fill = main), color = main) + 
  geom_bar(position="fill", stat="identity") +
  scale_fill_hue(l=50, h.start = 250) +
  scale_color_hue(l=50, h.start = 250) +
  labs(title = 'Major Rocket Launches by Decade', y = '# of Launches', x = '',
       caption = 'Major Rocket Launch is any launch >100 Apogee',
       fill = 'Area') + theme(legend.position="right") +
  theme(axis.line.y = 'none', axis.text.y = 'none') +
  theme_light()
```

We see from here that the US has led in every decade except the 1980s.  We also see an increase in Asia's presence Space Operations as we come to modern times.

Now we'll pick some of the more interesting decades and see how the individual organizations and nations compare with each other.

### The Eagle has landed - 1960s

Our first snapshot will be during the height of the space race.  In this period the USA launched 4,264 rockets compared to 2,178 from the Soviet Union and only 1,119 from the rest of the world.

```{r}
df_plt <- filter(df_decade, Decade == '1960')

bb <- ggplot(data = df_plt)

bb + geom_treemap(aes(area = launches, subgroup = StateName, fill = StateName)) + 
  geom_treemap_text(aes(area = launches, subgroup = StateName, label = AgencyName), alpha = .7,
                    color = 'white', reflow = T, place = 'bottomleft') +
  geom_treemap_subgroup_text(aes(area = launches, subgroup = StateName, label = AgencyName), 
                             alpha = 1, place = 'center') +
  theme(legend.position = 'none') +
  labs(title = 'Major Launches in the 1960s', 
       caption = paste('Total >100 Apogee Launches =',
                       as.character(sum(df_plt$launches))))  +
  scale_fill_hue(l=50, h.start = 270)
```

### Before the Fall - 1980s

The decade the US was really out-rocketed.  Here we see the Soviet Union launched 2058 rockets compared to the USA's 990.

```{r}
df_plt <- filter(df_decade, Decade == '1980')

bb <- ggplot(data = df_plt)

bb + geom_treemap(aes(area = launches, subgroup = StateName, fill = StateName)) + 
  geom_treemap_text(aes(area = launches, subgroup = StateName, label = AgencyName), alpha = .7,
                    color = 'white', reflow = T, place = 'bottomleft') +
  geom_treemap_subgroup_text(aes(area = launches, subgroup = StateName, label = AgencyName), 
                             alpha = 1, place = 'center') +
  theme(legend.position = 'none') +
  labs(title = 'Major Launches in the 2010s', 
       caption = paste('Total >100 Apogee Launches =',
                       as.character(sum(df_plt$launches))))  +
  scale_fill_hue(l=50, h.start = 270)
```

### Rise of Commercial Rocketry - 2010s

Finally, it's time to see what the modern landscape of rocketry.  Two interesting developments can be seen during this decade.  First, is the increase in commercial activity; with ULA (60 w/ LMA, 36 w/ Boeing),  ArianeSpace (84), SpaceX (69), and Blue Origin (9) leading the way.  Second, is the increased number of launches from Asia with 227 from China, 107 from India, 51 from Japan, and 46 from North Korea.

```{r}
df_plt <- filter(df_decade, Decade == '2010')

bb <- ggplot(data = df_plt)

bb + geom_treemap(aes(area = launches, subgroup = StateName, fill = StateName)) + 
  geom_treemap_text(aes(area = launches, subgroup = StateName, label = AgencyName), alpha = .7,
                    color = 'white', reflow = T, place = 'bottomleft') +
  geom_treemap_subgroup_text(aes(area = launches, subgroup = StateName, label = AgencyName), 
                             alpha = 1, place = 'center') +
  theme(legend.position = 'none') +
  labs(title = 'Major Launches in the 2010s', 
       caption = paste('Total >100 Apogee Launches =',
                       as.character(sum(df_plt$launches))))  +
  scale_fill_hue(l=50, h.start = 270)
```

## The Final Frontier

I've so far just scratched the surface of the information held in this dataset.  I look forward to making many more discoveries within it.

Thanks for the reading.